Understanding the Lomb-Scargle Periodogram

### 2017-02-01 PracticalLombScargle.tex
edits and rewordings

The Lomb-Scargle periodogram \citep{Lomb76, Scargle82}
is a well-known algorithm for detecting periodicity
in unevenly-sampled time-series, particularly within the astronomy community.
For example, consider the data shown in \fig{LINEAR-data}:
this is an irregularly-sampled timeseries showing a single object from the
LINEAR survey \citep{LINEAR1, LINEAR3}, with magnitude measured 280 times over
the course of five and a half years\footnote{
  Python code to reproduce \fig{LINEAR-data}, as well as all other figures
  in this manuscript, is available at 
  \url{http://github.com/jakevdp/PracticalLombScargle/}}.
By eye, it is clear that the brightness of the object varies in time with a range spanning approximately 0.8 magnitudes, but what is not immediately clear is that this variation is periodic in time.
The Lomb-Scargle periodogram is a method that allows efficient computation of a Fourier-like power spectrum from such unevenly-sampled data, resulting in an intuitive means of determining the period of oscillation.

@@

Often this is exactly how the Lomb-Scargle periodogram is presented: as a clean, well-defined procedure to detect the periodic component in an unevenly-sampled dataset.
In practice, however, there are a number of subtle issues that must be considered when applying a Lomb-Scargle analysis to real-world datasets, and I have found that these are rarely presented together at an introductory level.
Here are a few questions in particular that we might wish to ask about the results in \fig{LINEAR-power}:
\begin{enumerate}
  \item What is the source of the multiple peaks revealed by the Lomb-Scargle Periodogram?
  \item What is the largest frequency (i.e. Nyquist-like limit) that such an analysis is sensitive to?
  \item How should we choose the spacing of the frequency grid for our periodogram?
  \item What assumptions is the Lomb-Scargle periodogram making about the signal?
  \item How should we understand the uncertainty of the computed frequency?
\end{enumerate}
Quantitative answers to these sorts of questions are discussed in various textbooks and review papers, but I have not come across any single concise reference that gives a good intuition for how to think about such questions.
This paper seeks to fill that gap, and provide a practical, just-technical-enough guide to the effective use of the Lomb-Scargle method for periodic analysis.
This paper does not seek a complete or rigorous treatment of the mathematics involved, but rather seeks to develop the intuition of {\it how} to think about the issues involved.

@@

   \item[A time-shift imparts a phase in the Fourier transform.]
     Given a well-behaved function $g(t)$ we can use a transformation of
     variables to derive the following identity:
     \begin{equation}
       \mathcal{F}\{g(t - t_0)\} = \mathcal{F}\{g(t)\} e^{-2\pi i ft_0}
     \end{equation}
     The key observation here is that the time-shift does not change the
     amplitude of the resulting transform, but only the phase.
\end{description}
These properties taken together make the Fourier transform quite useful for the study of periodic signals.
The linearity of the transform means that any signal made up of a sum of periodic components will have a Fourier transform consisting of a sum of delta functions at those frequencies: that is, the Fourier transform is a {\it direct} measure of the periodic content of a function.

Further, if we compute the squared amplitude of the resulting transform, we can both do away with complex components and remove the phase imparted by any time offsets; this squared amplitude is usually known as the {\it Power Spectrum}:
\begin{equation}
  \mathcal{P}\{g\} \equiv \left|\mathcal{F}\{g\}\right|^2
  \eqlabel{power-spectrum}
\end{equation}
The power spectrum of a function is a positive real-valued function of $f$ that quantifies the contribution of each frequency $f$ to the total signal.

@@

  \item[The Fourier transform of a Dirac comb is a Dirac comb.]
    (See \fig{fourier-pairs}{\it d})
    \begin{equation}
      \mathcal{F}\{\III_T(t)\} = \frac{1}{T}\III_{1/T}(f)
    \end{equation}
    The Dirac comb, $\III_T(t)$, is an infinite sequence of Dirac delta functions placed at even intervals of size $T$:
    \begin{equation}
      \III_T(t) \equiv \sum_{n=-\infty}^\infty \delta(t - nT)
      \eqlabel{dirac-comb}
    \end{equation}
\end{description}
The key observation in each of these Fourier pairs is the reciprocity
of scales between a function and its Fourier transform:
a narrow function will have a broad transform, and vice versa.
More quantitatively, a function with a characteristic scale $T$ will in
general have a Fourier transform with characteristic scale of $1/T$.
Such reciprocity comes up in many diverse applications of Fourier analysis,
from wavelet transforms to music to quantum mechanics.
For our purposes, this feature will turn out to be quite important as we
push further in understanding the Lomb-Scargle Periodogram.

@@

Given this definition of a convolution, it can be shown that the Fourier transform of a convolution is the point-wise product of the Fourier transforms:
\begin{equation}
  \mathcal{F}\{f \ast g\} = \mathcal{F}\{f\} \cdot \mathcal{F}\{g\}
  \eqlabel{convolution-theorem}
\end{equation}
In practice, this is a much more efficient means of computing the convolution
than to directly solve at each time $t$ the integral over $\tau$ that appears
in \eq{convolution-definition}.
This result in \eq{convolution-theorem} is known as the
{\it convolution theorem}, and is illustrated in \fig{convolution-theorem}.
\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth]{fig05_Convolution_Theorem}
  \caption{Visualization of the convolution theorem (\eq{convolution-theorem}).
    The key is that the Fourier transform of
    a convolution is the pointwise product of the two Fourier transforms.
    In the right panels, the black and gray lines represent the real and
    imaginary part of the transform, respectively.
    \figlabel{convolution-theorem}}
\end{figure}
An important corrollary is that the Fourier transform of a product is a convolution of the two transforms:
\begin{equation}
  \mathcal{F}\{f \cdot g\} = \mathcal{F}\{f\} \ast \mathcal{F}\{g\}
  \eqlabel{convolution-theorem-inverse}
\end{equation}
We will see that these properties of the Fourier transform become essential
when thinking about frequency components of time-domain measurements.

@@

\todo{discuss data modes}

In a typical observational setting, we will be observing some small part of
a larger time-varying signal.
For example, if you have a very long signal that you measure for a finite
amount of time, your measurement is a point-wise product between the true
signal and a suitable rectangular function describing your measurement process.
If you have a continuous signal that you measure at regular intervals, your
measurement is a point-wise product between the true signal and an appropriate
Dirac comb describing your measurement process.
From what we learned about the convolution theorem in the previous section, it
is clear that {\it the measured Fourier transform will be the true transform
convolved with the transform of the obverving window}.
Symbolically, if the signal is $g(t)$ and the window is $W(t)$, the observed
function is
\begin{equation}
  g_{obs}(t) = g(t)W(t),
\end{equation}
and its Fourier transform is a convolution:
\begin{equation}
  \mathcal{F}\{g_{obs}\} = \mathcal{F}\{g\} \ast \mathcal{F}\{W\}.
\end{equation}
This has some interesting consequences, as we shall see.

@@

First, let's consider the case of observing a continuous periodic signal over
a limited span of time, as illustrated in \Fig{rectangular-window}.
Here we consider a periodic function observed only within the window $[-3, 3]$.
The observed signal in this case can be thought of as a pointwise product of an infinite periodic signal, and a rectangular window function.
Using \eq{convolution-theorem-inverse}, we see that the Fourier transform of
this function is equivalent to the convolution of the Fourier transforms of each component, which are a set of delta functions and a sinc function, respectively.
For a purely periodic signal, this convolution has the effect of broadening the delta functions representing the signal, turning each of them into a shifted sinc function.
Because of the inverse relationship between the width of the window and the width of its transform (see \Fig{fourier-pairs}), it follows that a wider observing window leads to proportionally less spread in the Fourier transform of the observed function.

@@

Interestingly, because the Fourier transform of a Dirac comb is another Dirac
comb, the effect of such an observing window is to create a long sequence
of aliases with a spacing of $1/T$.
With this in mind, we can be assured in this case that evaluating the
transform in the range $(2T)^{-1} \le f < (2T)^{-1}$ is sufficient to capture
all the available frequency information:
the signal outside that range is a sequence of identical aliases of
what lies in that range.
In fact, due to the nature of the aliasing, it is in fact sufficient to
compute the transform in {\it any} range of width $1/T$, and it is often
convenient to instead use the range $0 \le f < 1/T$.

@@

The example in \fig{comb-window-1} is somewhat of a best-case scenario, because
the true Fourier transform values lie entirely in a range of width $1/T$.
If we change the sampling rate so that this is no longer the case, we will
have a situation similar to that in \fig{comb-window-2}.
A sparser sampling of the signal leads to a narrower sampling in frequency,
and the signal transform no longer fits within the frequency window.
The result is that the true Fourier transform {\it cannot be recovered}
from the observed transform.

This implies that if we have a regularly-sampled function with a sampling
rate of $f_0 = 1/T$, we can only fully recover the freuqency information
if the signal is {\it band-limited} between frequencies $\pm f_0/2$.
This is one way to motivate the famous Nyquist sampling limit, which
approaches the question the other way around and says
that to fully represent the frequency content of a ``band-limited signal''
whose Fourier transform is zero outside the range $\pm B$,
we must sample the data with a rate of at least $f_{Ny} = 2B$.

@@

\subsubsection{The Discrete Fourier Transform}
For regularly-spaced observations defined by a Dirac comb, the delta functions
serve to collapse the Fourier integral into a sum over periodic components.
Suppose we have a true (infinitely long and continuous) signal $g(t)$, but
we observe it only at a regular grid with spacing $\Delta t$. In this case, our
observed signal is $g_{obs} = g(t) \III_T(t)$ and its Fourier transform is
\begin{equation}
  \hat{g}_{obs}(f) = \sum_{n=-\infty}^\infty g(n\Delta t) e^{-2\pi i f n \Delta t},
  \eqlabel{DFT-f-inf}
\end{equation}
which follows directly from \eq{FT-def} and \eq{dirac-comb}.

@@

You may have noticed we glossed over something here: the effect of switching
from an infinite number of samples to a finite number of samples.
In moving from \eq{DFT-f-inf} to \eq{DFT-f}, we have effectively applied
to our data a rectangular window function of width $NT$.
From the discussion accompanying \fig{rectangular-window}, we know what this does:
it gives us a Fourier transform convolved with a sinc function of width
$1 / (NT)$, resulting in the ``smearing'' of the Fourier transform signal
with this width.
Roughly speaking, then, any two Fourier transform values at frequencies within
$1/(NT)$ of each other will not be independent, and so we should space our
evaluations of the frequency with $\Delta f \ge 1/(NT)$.
Comparing to above, we see that this is {\it exactly the frequency spacing}
we arrived at from Nyquist-frequency arguments.

@@

With the discrete Fourier transform defined in \eqs{DFT-f}{DFT}, we can
apply the definition of the Fourier power spectrum from \eq{power-spectrum}
to compute the {\it Schuster Periodogram}, first proposed by \citet{Schuster98}:
\begin{equation}
  P_S(f) = \frac{1}{N}\left|\sum_{n=1}^N g_n e^{-2\pi i f t_n}\right|^2
  \eqlabel{schuster-periodogram}
\end{equation}
Apart from the $1/N$ scaling, this is precisely the Fourier power defined
in \eq{power-spectrum} for the expression in \eq{DFT-f}, and it follows that
for a uniformly-sampled function $g_n$, the Schuster periodogram captures
all of the relevant frequency information present in the data.

>>>>

The Lomb-Scargle periodogram \citep{Lomb76, Scargle82}
is a well-known algorithm for detecting periodicity in unevenly-sampled
time-series, and has seen particularly wide use within the astronomy community.
As an example of a typical application of this method,
consider the data shown in \fig{LINEAR-data}:
this is an irregularly-sampled timeseries showing a single object from the
LINEAR survey \citep{LINEAR1, LINEAR3}, with magnitude measured 280 times over
the course of five and a half years\footnote{
  Python code to reproduce \fig{LINEAR-data}, as well as all other figures
  in this manuscript, is available at 
  \url{http://github.com/jakevdp/PracticalLombScargle/}}.
By eye, it is clear that the brightness of the object varies in time with a
range spanning approximately 0.8 magnitudes, but what is not immediately clear
is that this variation is periodic in time.
The Lomb-Scargle periodogram is a method that allows efficient computation of
a Fourier-like power spectrum from such unevenly-sampled data, resulting in
an intuitive means of determining the period of oscillation.

@@

Often this is exactly how the Lomb-Scargle periodogram is presented: as a
clean, well-defined procedure to detect the periodic component in an
unevenly-sampled dataset.
In practice, however, there are a number of subtle issues that must be
considered when applying a Lomb-Scargle analysis to real-world datasets.
Here are a few questions in particular that we might wish to ask about
the results in \fig{LINEAR-power}:
\begin{enumerate}
  \item What is the source of the pattern of multiple peaks revealed by the
    Lomb-Scargle Periodogram?
  \item What is the largest frequency (i.e. Nyquist-like limit) that such
    an analysis is sensitive to?
  \item How should we choose the spacing of the frequency grid for our
    periodogram?
  \item What assumptions is the Lomb-Scargle periodogram making about the
    signal?
  \item How should we understand the uncertainty of the peak frequency?
\end{enumerate}
Quantitative answers to these sorts of questions are discussed in various
textbooks and review papers, but I have not come across any single concise
reference that gives a good intuition for how to think about such questions.
This paper seeks to fill that gap, and provide a practical,
just-technical-enough guide to the effective use of the Lomb-Scargle method
for detection and characterization of periodic signals.
This paper does not seek a complete or rigorous treatment of the mathematics
involved, but rather seeks to develop the intuition of {\it how} to think
about these questions.

@@

   \item[A time-shift imparts a phase in the Fourier transform.]
     Given a well-behaved function $g(t)$ we can use a transformation of
     variables to derive the following identity:
     \begin{equation}
       \mathcal{F}\{g(t - t_0)\} = \mathcal{F}\{g(t)\} e^{-2\pi i ft_0}
     \end{equation}
     The key observation here is that the time-shift does not change the
     amplitude of the resulting transform, but only the phase.
\end{description}
These properties taken together make the Fourier transform quite useful for the study of periodic signals.
The linearity of the transform means that any signal made up of a sum of
sinusoidal components will have a Fourier transform consisting of a sum
of delta functions marking the frequencies of those sinusoids: that is,
the Fourier transform is a {\it direct} measure of the periodic
content of a function.

Further, if we compute the squared amplitude of the resulting transform, we
can both do away with complex components and remove the phase imparted by any
time offsets; this squared amplitude is usually known as the
{\it Power Spectrum}:
\begin{equation}
  \mathcal{P}_g \equiv \left|\mathcal{F}\{g\}\right|^2
  \eqlabel{power-spectrum}
\end{equation}
The power spectrum of a function is a positive real-valued function of the
frequency $f$ that quantifies the contribution of each frequency $f$ to
the total signal.
Note that if $g$ is real-valued, it can be shown that
$\mathcal{P}_g(f) = \mathcal{P}_g(-f)$.

@@

  \item[The Fourier transform of a Dirac comb is a Dirac comb.]
    (See \fig{fourier-pairs}{\it d})
    \begin{equation}
      \mathcal{F}\{\III_T(t)\} = \frac{1}{T}\III_{1/T}(f)
    \end{equation}
    The Dirac comb, $\III_T(t)$, is an infinite sequence of Dirac delta functions placed at even intervals of size $T$:
    \begin{equation}
      \III_T(t) \equiv \sum_{n=-\infty}^\infty \delta(t - nT)
      \eqlabel{dirac-comb}
    \end{equation}
\end{description}
The key observation in each of these Fourier pairs is the reciprocity
of scales between a function and its Fourier transform:
a narrow function will have a broad transform, and vice versa.
More quantitatively, a function with a characteristic scale $T$ will in
general have a Fourier transform with characteristic scale of $1/T$.
Such reciprocity is central to many diverse applications of Fourier analysis,
from electrodynamics to music theory to quantum mechanics.
For our purposes, this property of the transform will turn out to be quite
important as we push further in understanding the Lomb-Scargle Periodogram.

@@

Given this definition of a convolution, it can be shown that the Fourier transform of a convolution is the point-wise product of the Fourier transforms of
the two functions:
\begin{equation}
  \mathcal{F}\{f \ast g\} = \mathcal{F}\{f\} \cdot \mathcal{F}\{g\}
  \eqlabel{convolution-theorem}
\end{equation}
In practice, this is a much more efficient means of computing the convolution
than to directly solve at each time $t$ the integral over $\tau$ that appears
in \eq{convolution-definition}.
This result in \eq{convolution-theorem} is known as the
{\it convolution theorem}, and is illustrated in \fig{convolution-theorem}.
\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth]{fig05_Convolution_Theorem}
  \caption{Visualization of the convolution theorem (\eq{convolution-theorem}).
    The key is that the Fourier transform of
    a convolution is the pointwise product of the two Fourier transforms.
    In the right panels, the black and gray lines represent the real and
    imaginary part of the transform, respectively.
    \figlabel{convolution-theorem}}
\end{figure}
An important corrollary is that the Fourier transform of a product is a convolution of the two transforms:
\begin{equation}
  \mathcal{F}\{f \cdot g\} = \mathcal{F}\{f\} \ast \mathcal{F}\{g\}
  \eqlabel{convolution-theorem-inverse}
\end{equation}
We will see that these properties of the Fourier transform become essential
when thinking about frequency components of time-domain measurements.

@@

Until now we have been discussing Fourier transforms of continuous signals
which are well-defined for all times $-\infty < t < \infty$.
Real-world temporal measurements of a signal, however, can only measure some
finite span of time, at some finite sampling rate.
In either case, the resulting data can be described by a point-wise product
of the true underlying continuous signal with a window function describing
the observation.
For example, a continuous signal measured over a finite duration is described
by a rectangular window function spanning the duration of the observation,
and a signal measured at regular intervals is described by a Dirac comb window
function marking those measurement times.\footnote{
  Throughout this paper we are concerned only with the mode of measurement
  common in optical astronomy; {i.e.} point observation. Other modes of
  observation, such as time-tag-events, binned event data, or
  time-to-spill events, are common in high-energy astronomy and other areas.
  We will not consider such event-driven data modes here, but it should be
  noted that there have been some interesting explorations of unified
  statistical treatments of all of the above modes of observation
  \citep[e.g.][]{Scargle98, Scargle2002}.
}

The Fourier transform we measure in these cases, then, is not the
transform of the continuous underlying function, but rather a 
{\it convolution of the true transform with the transform of the observing
window}.
Symbolically, if the signal is $g(t)$ and the window is $W(t)$, the observed
function is
\begin{equation}
  g_{obs}(t) = g(t)W(t),
\end{equation}
and its Fourier transform is a convolution of two transforms:
\begin{equation}
  \mathcal{F}\{g_{obs}\} = \mathcal{F}\{g\} \ast \mathcal{F}\{W\}.
\end{equation}
In the general case, such a convolution is not an invertible operation, meaning
that, except in special circumstances, it is not possible to exactly recover
the transform of the underlying signal.
This has some interesting consequences for the use and interpretation of
periodograms, as we shell see.

@@

First, let's consider the case of observing a continuous periodic signal over
a limited span of time, as illustrated in \Fig{rectangular-window}.
Here we consider a periodic function observed only within the window $-3<t<3$.
The observed signal in this case can be thought of as a pointwise product of
the underlying infinite periodic signal, and a rectangular window function.
Using \eq{convolution-theorem-inverse}, we see that the Fourier transform of
this function is equivalent to the convolution of the Fourier transforms of
each component, which are a set of delta functions and a sinc function,
respectively.
For a purely periodic signal, this convolution has the effect of broadening
the delta functions representing the signal, turning each of them into a
shifted sinc function.
Because of the inverse relationship between the width of the window and the
width of its transform (see \Fig{fourier-pairs}), it follows that a wider
observing window leads to proportionally less spread in the Fourier
transform of the observed function.

@@

Interestingly, because the Fourier transform of a Dirac comb is another Dirac
comb, the effect of such an observing window is to create a long sequence
of aliases of the underlying transform with a spacing of $1/T$.
With this in mind, we can be assured in this case that evaluating the
transform in the range $(2T)^{-1} \le f < (2T)^{-1}$ is sufficient to capture
all the available frequency information:
the signal outside that range is a sequence of identical aliases of
what lies in that range.
In fact, due to the nature of the aliasing, it is in fact sufficient to
compute the transform in {\it any} range of width $1/T$, and it is often
convenient to instead use the range $0 \le f < 1/T$.

@@

The example in \fig{comb-window-1} is somewhat of a best-case scenario, because
the true Fourier transform values are non-zero only within a range of width $1/T$.
If we change the sampling rate so that this is no longer the case, we will
have a situation similar to that in \fig{comb-window-2}.
A sparser sampling of the signal leads to a narrower sampling in frequency,
and the signal transform no longer fits within the frequency window.
The result is that the true Fourier transform {\it cannot be recovered}
from the observed transform.

This implies that if we have a regularly-sampled function with a sampling
rate of $f_0 = 1/T$, we can only fully recover the frequency information
if the signal is {\it band-limited} between frequencies $\pm f_0/2$.
This is one way to motivate the famous Nyquist sampling limit, which
approaches the question the other way around and says
that to fully represent the frequency content of a ``band-limited signal''
whose Fourier transform is zero outside the range $\pm B$,
we must sample the data with a rate of at least $f_{Ny} = 2B$.

@@

\subsubsection{The Discrete Fourier Transform}
For regularly-spaced observations defined by a Dirac comb, the delta functions
serve to collapse the Fourier integral into a sum over periodic components,
and in this way we can arrive at the common form of the discrete Fourier
transform.
Suppose we have a true (infinitely long and continuous) signal $g(t)$, but
we observe it only at a regular grid with spacing $\Delta t$. In this case, our
observed signal is $g_{obs} = g(t) \III_T(t)$ and its Fourier transform is
\begin{equation}
  \hat{g}_{obs}(f) = \sum_{n=-\infty}^\infty g(n\Delta t) e^{-2\pi i f n \Delta t},
  \eqlabel{DFT-f-inf}
\end{equation}
which follows directly from \eq{FT-def} and \eq{dirac-comb}.

@@

You may have noticed we glossed over something here: the effect of switching
from an infinite number of samples to a finite number of samples.
In moving from \eq{DFT-f-inf} to \eq{DFT-f}, we have effectively applied
to our data a rectangular window function of width $N\Delta t$.
From the discussion accompanying \fig{rectangular-window}, we know what this
does: it gives us a Fourier transform convolved with a sinc function of width
$1 / (N\Delta t)$, resulting in the ``smearing'' of the Fourier transform signal
with this width.
Roughly speaking, then, any two Fourier transform values at frequencies within
$1/(N\Delta t)$ of each other will not be independent, and so we should space
our evaluations of the frequency with $\Delta f \ge 1/(N\Delta t)$.
Comparing to above, we see that this is {\it exactly the frequency spacing}
we arrived at from Nyquist-frequency arguments.

@@

With the discrete Fourier transform defined in \eqs{DFT-f}{DFT}, we can
apply the definition of the Fourier power spectrum from \eq{power-spectrum}
to compute the {\it Schuster Periodogram}, first proposed by \citet{Schuster98}:
\begin{equation}
  P_S(f) = \frac{1}{N}\left|\sum_{n=1}^N g_n e^{-2\pi i f t_n}\right|^2
  \eqlabel{schuster-periodogram}
\end{equation}
Apart from the $1/N$ scaling, this sum is precisely the Fourier power
spectrum in \eq{power-spectrum}, computed for a continuous signal with
uniform sampling.
It follows that, in the uniform sampling case the Schuster periodogram captures
all of the relevant frequency information present in the data.
